---
title: "Lab 3"
output:
  pdf_document: default
  html_document: default
---

```{r}
library(tidyverse)
library(plotrix)
density <- read_csv("../data_raw/Lab3_Density.csv")
time_density <- read_csv("../data_raw/Lab3_densitytime.csv")
time_frequency <- read_csv("../data_raw/Lab3_frequency_time.csv")
```

# 1.) Calculate the mean and standard error of mesquite density for each plot size and make a graph of the comparison. (Be sure to convert density)

```{r}
#longer data tables
density_data <- density %>%
  pivot_longer(
    cols = c(`50x50cm`, `1x1m`, `2.5x2.5m`, `5x5m`, `10x10m`),
    names_to = "plot_size",
    values_to = "count"
  ) %>%
  mutate(
    # Set the area in m^2 for each plot size
    area_m2 = case_when(
      plot_size == "50x50cm"  ~ 0.25,
      plot_size == "1x1m"     ~ 1.00,
      plot_size == "2.5x2.5m" ~ 6.25,
      plot_size == "5x5m"     ~ 25.00,
      plot_size == "10x10m"   ~ 100.00), 
    #convert density
    density = count/area_m2,
    #set as factor to order smallest to largest quadrat and order x-axis titles
    plot_size = factor(plot_size, 
      levels = c("50x50cm", "1x1m", "2.5x2.5m", "5x5m", "10x10m")
    )
  )

#standard error and standard deviation
summary_density <- density_data %>%
  group_by(plot_size) %>%
  summarise(
    mean_density = mean(density),
    se_density   = sd(density) / sqrt(n())
  )

print(summary_density)
    
#plot summary stats
ggplot(summary_density, aes(x = plot_size, y = mean_density, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  geom_errorbar(
    aes(ymin = mean_density - se_density, ymax = mean_density + se_density,
    color = "Standard Error ±1 SE"),
width = 0.15, linewidth = 0.8
  ) +
  scale_color_manual(name = "Uncertainty", values = c("Standard Error ±1 SE" = "orange")) +
  labs(
    title = "Mesquite Density Mean Across Plot Sizes",
    x = "Plot Size",
    y = "Mesquite Density plants/m^2")

```

# 2.) Calculate the average time to sample each plot size.

```{r}

time_density_longer <- time_density %>%
  pivot_longer(
    cols = c(`50x50cm`, `1x1m`, `2.5x2.5m`, `5x5m`, `10x10m`),
    names_to = "plot_size",
    values_to = "count")

time_density_longer <- time_density_longer %>%
  mutate(
    plot_size = factor(
      plot_size, 
      levels = c("50x50cm", "1x1m", "2.5x2.5m", "5x5m", "10x10m") ) )

summary_time_density <- time_density_longer %>%
  group_by(plot_size) %>%
  summarise(mean_time = mean(count))

time_frequency_longer <- time_frequency %>% 
  pivot_longer(
    cols = c(`50x50cm`, `1x1m`, `2.5x2.5m`, `5x5m`, `10x10m`),
    names_to = "plot_size",
    values_to = c("count"))

time_frequency_longer <- time_frequency_longer %>% 
  mutate(
    plot_size = factor( plot_size,
    levels = c("50x50cm", "1x1m", "2.5x2.5m", "5x5m", "10x10m")
  ) )

summary_time_frequency <- time_frequency_longer %>% 
  group_by(plot_size) %>% 
  summarise(mean_time = mean(count))

# 1. Combine both summaries and tag the sampling method
summary_combined <- bind_rows(
  summary_time_density %>% mutate(method = "Density"),
  summary_time_frequency %>% mutate(method = "Frequency")
)

# 2. Plot side-by-side bars using position_dodge()
ggplot(summary_combined, aes(x = plot_size, y = mean_time, fill = method)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_text(
    aes(label = round(mean_time, 1)),
    position = position_dodge(0.8),
    vjust = -0.5,
    size = 3.5
  ) +
  scale_fill_manual(
    name = "Sampling Method",
    values = c("Density" = "blue", "Frequency" = "orange")
  ) +
  labs(
    title = "Mean Sampling Time Comparison Across Plot Sizes",
    x = "Plot Size",
    y = "Mean Time (seconds)"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position = "top"
  )
  
```


# 3.) In at least a paragraph, answer which is the most precise (least variable) plot size and why.

# 4.) In at least a paragraph, answer which is the most efficient plot size and why.

# 5.) Converting the density to frequency, what plot size worked well for frequency?

# 6.) Theoretically frequency (p in %) and density (d) are related. Some suggest that the conversion from frequency to density is d = -ln(1-p/100). Calculate the estimated density for the 5x5m quadrat frequency data. (Convert to plants per m^2) Does the answer come out the same or different from the actual density answer?

# 7.) Write at least a paragraph commenting about this exercise.

# 8.) For the 2.5x2.5m plot size, calculate teh sample size to produce, with 95% probability, a sample mean within 10% of the true population mean. What does this say about our ability to detect changes with the sample size of 20?